---
title: "Population Description"
author: "David Hughes"
date: "3/16/2022"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#### LIBRARIES
library(tidyverse)
library(gtsummary)
#### Parameters
source("load.R")
```


## Population Description

```{r}
meta_data = study_data$meta_data

###
cat(paste0("\n How many individuals do not have an assigned sex?"))
sum( is.na(meta_data$gender_online) )
cat(paste0("\n Males and Females (v)?"))
table( meta_data$gender_online )
cat(paste0("\n Males and Females (v) in the imputed data?"))
table( meta_data$imputed_sex )

```


```{r}
temp = meta_data |> 
  # filter(exclude_subject == FALSE) |>
  # dplyr::select(gender_online, age_online, height, weight_online, BMI_online, smoking_NEC, house_hold_monthl_net_income )
  dplyr::select(imputed_sex, imputed_age, imputed_height, 
                imputed_weight_online, imputed_BMI_online, smoking_NEC, house_hold_monthl_net_income )

temp$imputed_age = as.numeric(temp$imputed_age)
temp$imputed_sex = gsub("m","male",temp$imputed_sex )
temp$imputed_sex = gsub("v","female",temp$imputed_sex )


colnames(temp) = c("sex","age","height","weight","BMI","smoking","household income")


Q2.5 <- \(x) quantile(x, probs = 0.025, type = 7)
Q97.5 <- \(x) quantile(x, probs = 0.975, type = 7)


tbl = temp |> 
  tbl_summary( by = sex, missing = "no",
               statistic = all_continuous() ~ "{mean} ({Q2.5}, {Q97.5})",
               ) |>
  add_overall() |>
  # add_p() |> 
  add_n() |>
  bold_labels()



#f = paste0(project_results_dir, "tables/pop_description_20240917.docx")
f = paste0(project_results_dir, "tables/pop_description_20242511_v2.png")
tbl |> 
  as_gt() |> 
  gt::gtsave(filename = f)


f = paste0(project_results_dir, "tables/pop_description_20242511_v2.txt")
write.table(as.tibble(tbl), f, sep = "\t", row.names = FALSE, quote = FALSE)
f = paste0(project_results_dir, "tables/pop_description_20242511_v3.txt")
write.table( tbl |> as_kable(), f, sep = "\t", row.names = FALSE, quote = FALSE)

tbl # |> as_kable() |> kableExtra::kable_classic_2()
```



## is BMI different between males and females?

```{r}
bmi = temp$BMI
m = which(temp$sex == "male")
f = which(temp$sex == "female")
t.test(bmi[m], bmi[f])$p.value
```

## What proportion of the sample population has a HEALTHY, OVERWEIGHT, OBESE BMI state ?

```{r}
x = na.omit(bmi)
## Healthy
sum( x < 25 )/length(x)
## Overweight
sum( x >= 25 & x < 30 )/length(x)
## Obese
sum( x > 30 )/length(x)
```

## males

```{r}
x = na.omit(bmi[m])
## Healthy
sum( x < 25 )/length(x)
## Overweight
sum( x >= 25 & x < 30 )/length(x)
## Obese
sum( x > 30 )/length(x)
```


## females

```{r}
x = na.omit(bmi[f])
## Healthy
sum( x < 25 )/length(x)
## Overweight
sum( x >= 25 & x < 30 )/length(x)
## Obese
sum( x > 30 )/length(x)
```


# Variance explained in exposure by instrument

## Yengo BMI and Height

```{r}
d = data.frame( id = study_data$meta_data$vdp_ids ,
                bmi = study_data$meta_data$imputed_BMI_online, 
                height = study_data$meta_data$imputed_height,
                whr = study_data$meta_data$imputed_WHR,
                whradjbmi = study_data$meta_data$imputed_WHRadjBMI,
                sex = study_data$meta_data$imputed_sex, 
                age = study_data$meta_data$imputed_age,
                
                yengo = study_data$meta_data$yengo_bmi_primary,
                pulit = study_data$meta_data$pulit_bmi,
                pulit_f = study_data$meta_data$pulit_bmi_females,
                pulit_m = study_data$meta_data$pulit_bmi_males,
                yengo_h = study_data$meta_data$yengo_height_primary,
                pulit_whr = study_data$meta_data$pulit_whr,
                pulit_whr_f = study_data$meta_data$pulit_whr_females,
                pulit_whr_m = study_data$meta_data$pulit_whr_males,
                pulit_whradjbmi = study_data$meta_data$pulit_whradjbmi,
                pulit_whradjbmi_f = study_data$meta_data$pulit_whradjbmi_females,
                pulit_whradjbmi_m = study_data$meta_data$pulit_whradjbmi_males,
                pc1 = study_data$meta_data$PC1,
                pc2 = study_data$meta_data$PC2,
                pc3 = study_data$meta_data$PC3,
                pc4 = study_data$meta_data$PC4 )
rownames(d) = d$id

```


```{r}

cat(paste0("Yengo BMI in both sexes\n"))
fit = lm( bmi ~ yengo, data = d )
s = summary(fit); s$r.squared

cat(paste0("Pulit BMI in both sexes\n"))
fit = lm( bmi ~ pulit, data = d )
s = summary(fit); s$r.squared

cat(paste0("Height in both sexes\n"))
fit = lm( height ~ yengo_h, data = d )
s = summary(fit); s$r.squared

cat(paste0("Pulit WHR in both sexes\n"))
fit = lm( whr ~ pulit_whr, data = d )
s = summary(fit); s$r.squared

cat(paste0("Pulit WHRadjBMI in both sexes\n"))
fit = lm( whradjbmi ~ pulit_whradjbmi, data = d )
s = summary(fit); s$r.squared

```

##  BMI multivariable eta-squared

```{r}
fit = lm( bmi ~ sex + age + pulit, data = d )
a = anova(fit)
ss = a[,2]; names(ss) = rownames(a)
## eta-squared
ss/sum(ss)

```


## Pulit BMI by sex

```{r}

f = which(d$sex == "v")
m = which(d$sex == "m")

###
cat(paste0("BMI in both sexes\n"))
fit = lm( bmi ~ pulit, data = d )
s = summary(fit); s$r.squared

###
cat(paste0("\nBMI in females\n"))
fit = lm( bmi ~ pulit, data = d[f,] )
s = summary(fit); s$r.squared


cat(paste0("\nfemale_BMI in females\n"))
fit = lm( bmi ~ pulit_f, data = d[f,] )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in males\n"))
fit = lm( bmi ~ pulit, data = d[m,] )
s = summary(fit); s$r.squared


cat(paste0("\nmale BMI in males\n"))
fit = lm( bmi ~ pulit_m, data = d[m,] )
s = summary(fit); s$r.squared

```

## Pulit WHR

```{r}
cat(paste0("BMI in both sexes\n"))
fit = lm( whr ~ pulit_whr, data = d )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in females\n"))
fit = lm( whr ~ pulit_whr_f, data = d[f,] )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in males\n"))
fit = lm( whr ~ pulit_whr_m, data = d[m,] )
s = summary(fit); s$r.squared


```


## Pulit WHRadjBMI

```{r}
cat(paste0("BMI in both sexes\n"))
fit = lm( whradjbmi ~ pulit_whradjbmi, data = d )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in females\n"))
fit = lm( whradjbmi ~ pulit_whradjbmi_f, data = d[f,] )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in males\n"))
fit = lm( whradjbmi ~ pulit_whradjbmi_m, data = d[m,] )
s = summary(fit); s$r.squared

```

## Correlation between the Yengo and Pulit BMI PGS score

```{r}
cor.test(d$yengo, d$pulit)
```

```{r, figwidth = 6, fig.height = 6}
study_data$micro_data %>% ggplot(aes(x = MDS1, y = MDS2, color = EnterotypeClass)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_color_brewer(palette = "Set1") +
  theme_bw()
  
```

