---
title: "Define Study Microbial Traits"
author: "David Hughes"
date: "3/16/2022"
output: html_document
---

How shall we handle the taxanomic count filtering and the building of the MTs (microbial traits) ?


```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#### LIBRARIES
library(tidyverse)

#### Parameters
source("parameters/pfile.sh")
source("load.R")
```

## Define initial study microbial traits data

```{r}

############################################
## Limit the microbial traits to those with
##  -- 1. at least 20 non-zero individuals
##  ***  NO STEP 2 FILTER  ***
##  -- 2. a zero-truncated mean count of 25 OR a minimum of 500 (5%) reads in at least 1 individual
############################################
## all taxa in data set
all_mt_taxa = colnames(study_data$mt_data)[8:505]

## Build the MT data
MTdata = build_mts( wdata = study_data$mt_data[, all_mt_taxa], 
                      PA_min_obs = 250,   ## The minimum number of PRESENT of ABSCENT Observations to build a PA trait
                      trunc_AB_present_count = 1000, ## The minimum number of present observations to build a trunc_AB
                      trunc_AB_absence_count = 250, ## The minimum number of absent (0's) observations to build a trunc_AB
                      trunc_AB_mean = 20,  ## The minimum zero-truncated mean to build a trunc_AB
                      AB_zero_present_proportion = 0.75, ## The maximum proportion of zero data to build an AB trait
                      AB_mean_of_zero_trunc_data = 20 ## The minimum zero-truncated mean to build an AB trait
)
MTdata = as.data.frame(MTdata)


```


## Correlation Structure

```{r , warning=FALSE, message=FALSE, error=FALSE}
## Build a Spearman's correlation matrix
Cmat = cor(MTdata, method = "sp", use = "pairwise.complete.obs")

```

```{r, fig.width = 20, fig.height = 20}
M = Cmat
M[is.na(M)] = 0

corrplot::corrplot(M, method = 'color', order = 'hclust',  diag = FALSE, tl.cex = 0.3, tl.col = 'black')
```


```{r}
f = paste0(project_results_dir, "figures/mt_corplot.pdf")
pdf(file = f, width = 20, height = 20)
corrplot::corrplot(M, method = 'color', order = 'hclust',  diag = FALSE, tl.cex = 0.3, tl.col = 'black')
dev.off()

```

## Identify the redundant, higher order taxa (MTs)

```{r}
Dmat = as.dist(1-abs(M))
nj = hclust(Dmat)
k = cutree(nj, h = 0.01)
nrow(M)
length(unique(k))
```


```{r}
TAXAORDER = c("K","P","C","O","F","G")

## Identify K group with more than 1 MT
K_groups = names(  which( table(k) > 1 ) )
## Extract those MTs that ARE NOT the lowest taxanomic order
taxa_2_remove = c()
##
for(i in K_groups){
  ## extract MT names
  mt_names = names( which(k == i) )
  ## screen for trunc_AB and AB traits
  if( length( grep("trunc_AB", mt_names) ) > 0 & length( grep("trunc_AB", mt_names) ) != length(mt_names)  ){
    ## take the trunc_AB taxa
    q = grep("trunc_AB", mt_names)
    trunc_AB_mts = mt_names[q]
    AB_mts = mt_names[-q]
    ##########################
    ## Process the trunc_AB taxa
    ##########################
    if(length(trunc_AB_mts) > 1){
      to = substring(trunc_AB_mts, 1,1)
      m = na.omit( match(TAXAORDER, to) )
      r = m[-length(m)]
      taxa_2_remove = c(taxa_2_remove, trunc_AB_mts[r] )
    }
    ##########################
    ## Process the AB taxa
    ##########################
    if(length(AB_mts) > 1){
      to = substring(AB_mts, 1,1)
      m = na.omit( match(TAXAORDER, to) )
      r = m[-length(m)]
      taxa_2_remove = c(taxa_2_remove, AB_mts[r] )
    }
  
    ### If the mt_names do not include both trunc_AB and AB taxa names
    ### Process here:
  } else {
    ## extract taxa order
    to = substring(mt_names, 1,1)
    m = na.omit( match(TAXAORDER, to) )
    r = m[-length(m)]
    taxa_2_remove = c(taxa_2_remove, mt_names[r] )
  }
  
  
}
```

## how many redundant MTs to remove ?

```{r}
length(taxa_2_remove)
```
## A manual look at reducndant MTs, as defined by the tree cut above.

```{r}
for(i in K_groups){
  ## extract MT names
  mt_names = names( which(k == i) )
  cat( paste0("k = ", i, "; mts = ", paste(mt_names, collapse = " "), "\n\n" ) )
}
```


## Remove the redundant taxa from the MTdata data frame

```{r}
w = which(colnames(MTdata) %in% taxa_2_remove)
study_mt_data = MTdata[,-w]
```


```{r}
## Define a STUDY MICROBIOAL TRAIT Data Frame
study_data$study_mt_data = cbind( study_data$mt_data[, c("enterotype", "EnterotypeClass", "MDS1", "MDS2", 
                                             "Div_NumberGenera","Div_Shannon","Div_Chao1")] , study_mt_data)
```


```{r}
#################################
## RE-WRITE The project data frame
#################################
n = paste0( project_data_dir , "study_data_v1.1.Rdata")
save(study_data, file = n)

```





## Correlation Structure of STUDY MT DATA set

```{r , warning=FALSE, message=FALSE, error=FALSE}
## Build a Spearman's correlation matrix
Cmat = cor(study_data$study_mt_data[,-c(1:2)], method = "sp", use = "pairwise.complete.obs")

```

```{r, fig.width = 20, fig.height = 20}
M = Cmat
M[is.na(M)] = 0

corrplot::corrplot(M, method = 'color', order = 'hclust',  diag = FALSE, tl.cex = 0.3, tl.col = 'black')
```


```{r}
f = paste0(project_results_dir, "figures/mt_corplot_no_fullyredundant_taxa.pdf")
pdf(file = f, width = 20, height = 20)
corrplot::corrplot(M, method = 'color', order = 'hclust',  diag = FALSE, tl.cex = 0.3, tl.col = 'black')
dev.off()

```