---
title: "Identifying confounders"
author: "David Hughes"
date: "3/16/2022"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ivreg)

source("parameters/pfile.sh")
source("load.R")

```



```{r}
## Define the MetaData table
meta_data = study_data$meta_data

## Define the Microbial Trait Data
mt_data = study_data$study_micro_data
primary_traits = study_data$primary_traits
primary_traits = c("Enterotype_Bact1","Enterotype_Bact2","Enterotype_Prev","Enterotype_Rum",
                   "Div_NumberGenera","Div_Shannon","Div_Chao1",
                   "MDS1","MDS2",
                   primary_traits)

## Define the batch_data table
batch_data = study_data$batch_data
for(i in 1:8){ batch_data[,i] = as.factor(batch_data[,i])  }
```

## Build the working data frame

```{r}
## working data
mydata = cbind( meta_data, batch_data, mt_data )
## we will be using imputed BMI as the default BMI
w = which(colnames(mydata) == "imputed_BMI_online"); colnames(mydata)[w] = "BMI"
## we will be using imputed sex as the default sex
w = which(colnames(mydata) == "imputed_sex"); colnames(mydata)[w] = "sex"
mydata$sex[mydata$sex == "m"] = "male"
mydata$sex[mydata$sex == "v"] = "female"
mydata$sex = as.factor(mydata$sex)
## we will be using imputed age as the default age
w = which(colnames(mydata) == "imputed_age"); colnames(mydata)[w] = "age"
mydata$age = as.numeric(mydata$age)

## alter colnames with a "|" as a function below does not like "|" 
colnames(mydata) = gsub("\\|","__",colnames(mydata))


```


## Define covariables to test for association with Microbial Traits

```{r}
## COVARS
cvs = c( "imputed_hip_circumference",
         "imputed_waist_circumference",
         "imputed_WHR",
         "imputed_WHRadjBMI",
         "imputed_height",
         "fasting",
        colnames(mydata)[c(27:122,136,139, 141:149)]
        )
## For each variable tha was recoded remove the original
w = grep("\\.r", cvs)
w = cvs[w]
w = gsub("\\.r","",w)
r = which(cvs %in% w)
cvs = cvs[-r]
## turn chr into factor
for(i in cvs){
  if( is.character(mydata[,i]) ){
    mydata[,i] = as.factor(mydata[,i])
  }
}
## remove the exclusion variables
w = grep("exclu", cvs)
cvs = cvs[-w]

## remove diabetes type as it is limited to only those with diabetes
w = grep("diabetes_type", cvs)
cvs = cvs[-w]

## make factors
x = c("have_high_blood_cholesterol.r","have_high_blood_pressure.r","diabetes_type","have_diabetes.r",
      "limited_fat_diet","fiber_rich_diet", "weight_loss_diet", "gluten_free_diet", "lactose_free_diet", 
       "is_followingadiet", "is_vegan", "is_vegetarian")
for(i in x){
  mydata[, i] = factor(mydata[, i], levels = c(0,1))
}

```


### How many indpendent covariables are there?

```{r}
## define the temporary working data set
temp = mydata[, cvs]

## Define an empty matrix
cvs_CorMat = matrix(0, nrow = length(cvs), ncol = length(cvs), dimnames = list(cvs, cvs))

for(i in 1:length(cvs)){
  for(j in 1:length(cvs)){
    # cat(paste0("running ", i , " and ", j, "\n"))
    x = temp[,i]
    y = temp[, j]
    if( class(x) == "factor" & class(y) == "factor" ){
      o = moosefun::factor_on_factor(x, y)
      r = o["Crammers_V"]
    } 
    if( class(x) == "numeric" & class(y) == "numeric" ){
      o = moosefun::numeric_on_numeric(x, y) 
      r = o["rho"]
    }
    if( class(x) == "numeric" & class(y) == "factor" ){
      o = moosefun::factor_on_numeric(cat_values = y, num_values = x)
      r = o["rho"]
    }
    if( class(x) == "factor" & class(y) == "numeric" ){
      o = moosefun::factor_on_numeric(cat_values = x, num_values = y)
      r = o["rho"]
    }
    cvs_CorMat[i,j] = r
    
  }
}

```

```{r}
x = cvs_CorMat
x[is.na(x)] = 0
pca = prcomp(x, center = FALSE, scale = FALSE)
sum(summary(pca)[[6]][3,] <0.95) + 1

```


```{r, fig.width = 20, fig.height = 10}
dmat = 1 - abs(cvs_CorMat) 
dmat[is.na(dmat)] = 1
dmat = round(dmat, digits = 3)
dmat = as.dist(dmat)
nj = hclust( dmat, method = "complete")
k = cutree(nj, h = 0.8)
# length( table(k) )
plot(nj, hang = -1, main = paste0("Possible confounders for FGFP BMI->MT MR; Me = ", length( table(k) ), 
                                  "\nNumber of PCs to explaine 95% of the variance  = 43") )
abline(h = 0.8, col = "red", lty = 2, lwd = 2)
```

```{r,fig.width = 20, fig.height = 10}
f = paste0(project_results_dir, "figures/confounder_dendrogram.pdf")

pdf(file = f,width = 20, height = 9)
plot(nj, hang = -1, main = paste0("Possible confounders for FGFP BMI->MT MR; Me = ", length( table(k) ), 
                                  "\nNumber of PCs to explaine 95% of the variance  = 43") )
abline(h = 0.8, col = "red", lty = 2, lwd = 2)
dev.off()
```


```{r, fig.width = 20, fig.height = 10}
dmat = 1- cvs_CorMat
dmat[is.na(dmat)] = 1
dmat = round(dmat, digits = 3)
dmat = as.dist(dmat)
nj = hclust( dmat, method = "complete")
k = cutree(nj, h = 0.8)
# length( table(k) )
plot(nj, hang = -1, main = paste0("Possible confounders for FGFP BMI->MT MR; Me = ", length( table(k) ), 
                                  "\nNumber of PCs to explaine 95% of the variance  = 43") )
abline(h = 0.8, col = "red", lty = 2, lwd = 2)
```


## run linear models and anova

### AB traits

```{r, warning=FALSE, error=FALSE, message=FALSE}
## AB TRAITS
w = grep("_AB", colnames(mydata) )
q = grep("_RA", colnames(mydata) )
mts = colnames(mydata)[c(w,q)]
## add diversity traits to AB run
div = c("MDS1","MDS2","Div_NumberGenera","Div_Shannon","Div_Chao1")
mts = c(div, mts)
      
## ANALYSIS LOOP
BatchCors = c()
for(cv in cvs){
  cat(paste0("Processing cv ",cv ,"\n" ))
  out = t( sapply(mts, function(mt){
    lmfit_v3( wdata = mydata,
           outcome = mt,
           exposure = cv,
           covariates = c("sex","age"),
           family = "gaussian",
           weights = NA,
           rnt_outcome = TRUE)
  }) )
  out = as.data.frame(out)
  for(i in 3:ncol(out)){
    out[,i] = as.numeric(out[,i])
  }
  BatchCors = rbind(BatchCors, out)
  
}

# temp_BatchCors = BatchCors
```


### PA traits

```{r, warning=FALSE, error=FALSE, message=FALSE}
## PA TRAITS
w = grep("_PA", colnames(mydata) )
mts = colnames(mydata)[w]

## ANALYSIS LOOP
for(cv in cvs){
  cat(paste0("Processing cv ",cv ,"\n" ))
  out = t( sapply(mts, function(mt){
    lmfit_v3( wdata = mydata,
           outcome = mt,
           exposure = cv,
           covariates = c("sex","age"),
           family = "binomial",
           weights = NA,
           rnt_outcome = FALSE)
  }) )
  out = as.data.frame(out)
  for(i in 3:ncol(out)){
    out[,i] = as.numeric(out[,i])
  }
  ## ADDING ONTO THE AB modeling above
  BatchCors = rbind(BatchCors, out)
}

# temp_BatchCors = BatchCors
```


```{r}
BatchCors$exposure = unlist(BatchCors$exposure)
BatchCors$outcome = unlist(BatchCors$outcome)
```

## Write estimates to file

```{r}
# f = paste0(project_results_dir, "tables/possible_confounders_on_MTs_20240918.txt")
f = paste0(project_results_dir, "tables/possible_confounders_on_MTs_20241125.txt")
# write.table(BatchCors, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = TRUE)

# BatchCors = read.table(f, header = TRUE, sep = "\t", as.is = TRUE)
```

## Each of the 88 tested variables associate with how many MTs ?

```{r}
poss_confounders_ss = BatchCors |> filter(outcome %in% primary_traits) |> group_by(exposure) %>% 
  summarize( sig_count = sum(devexp_exposure_P < 0.05/33 ), 
             avg_dev = mean(devexp_by_exposure),
             min_dev = min(devexp_by_exposure),
             max_dev = max(devexp_by_exposure) ) %>% 
  arrange( desc(sig_count) )

poss_confounders_ss %>% knitr::kable() %>% kableExtra::kable_classic()
```


```{r}
cor.test(mydata$bristol_stool_score, mydata$defecation_days_count)
cor.test(mydata$bristol_stool_score, mydata$average_defecations_per_day.r)
cor.test(mydata$defecation_days_count, mydata$average_defecations_per_day.r)
cor.test(mydata$bristol_stool_score , mydata$previous_relief.r )
```



```{r}
cat(paste0("how many possible confounders associates with at least one MT ?\n"))
sum(poss_confounders_ss$sig_count>0, na.rm = TRUE)
cat(paste0("how many possible confounders associates with NO MT ?\n"))
sum(poss_confounders_ss$sig_count==0, na.rm = TRUE)
```
## Each MT is influenced by how many of the 88 tested variables?

```{r}
MT_confounders_ss = BatchCors %>% group_by(outcome) %>% 
  summarize( sig_count = sum(devexp_exposure_P < 0.05/43, na.rm = TRUE), 
             avg_dev = mean(devexp_by_exposure),
             min_dev = min(devexp_by_exposure),
             max_dev = max(devexp_by_exposure) ) %>% 
  arrange( desc(sig_count) )

MT_confounders_ss %>% knitr::kable() %>% kableExtra::kable_classic()
```


## dendrogram for MTs

```{r}
mts = unique(BatchCors$outcome)
mt_mat = cor(mydata[, mts])
mt_mat[is.na(mt_mat)] = 0
dmat = as.dist( 1-abs(mt_mat) )
nj_out = hclust(dmat, method = "complete")
```


```{r}
x = BatchCors %>% group_by(exposure) %>% 
  summarize(meanP = mean( -log10(devexp_exposure_P) )) %>% 
  arrange( desc(meanP) ) %>%
  dplyr::select(exposure)
sort_exposure_by_mean_P = x$exposure

############                                                
           
x = BatchCors %>% group_by(outcome) %>% 
  summarize(meanP = mean( -log10(devexp_exposure_P) )) %>% 
  arrange( desc(meanP) ) %>%
  dplyr::select(outcome)
sort_outcome_by_mean_P = x$outcome

```

```{r}
### ORDER THE EXPOSURES
# BatchCors$exposure = factor(BatchCors$exposure, levels = sort_exposure_by_mean_P)
BatchCors$exposure = factor(BatchCors$exposure, levels = nj$labels[nj$order] )

### ORDER THE OUTCOMES
BatchCors$outcome = gsub("__"," | ",BatchCors$outcome)
sort_outcome_by_mean_P = nj_out$labels[nj_out$order]
sort_outcome_by_mean_P = gsub("__"," | ", sort_outcome_by_mean_P )

BatchCors$outcome = factor(BatchCors$outcome, levels = sort_outcome_by_mean_P)

```


```{r, fig.width = 15, fig.height = 20}
plot = BatchCors |> filter(exposure != "time_of_sampling") |> 
  ggplot(aes(x = exposure, y = outcome)) +
  geom_raster(aes(fill = -log10(devexp_exposure_P))) +
  # scale_fill_viridis_c(direction = -1, option = "D") +
  # scale_fill_gradient2(low = "white", mid = "green4", high = "grey10", midpoint = 40) +
  # scale_fill_gradient2(low = "white", mid = "grey50", high = "black", midpoint = 30) +
  scale_fill_gradient2(low = "white", mid = "dodgerblue", high = "black", midpoint = 30) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position="top") +
  theme(axis.text.y = element_text(size = 2.5))

# f = paste0(project_results_dir, "figures/MT_Covariable_Assoction_TilePlot.pdf")
# pdf(file = f, width = 15, height = 35)
# plot
# dev.off()

plot
```


## Association of Covariables on BMI

```{r}
BMIcor = t( sapply(cvs, function(cv){
    lmfit_v3( wdata = mydata,
           outcome = "BMI",
           exposure = cv,
           covariates = c("sex","age"),
           family = "gaussian",
           weights = NA,
           rnt_outcome = FALSE)
  }) )
###
BMIcor = as.data.frame(BMIcor)
  for(i in 3:ncol(BMIcor)){
    BMIcor[,i] = as.numeric(BMIcor[,i])
  }
BMIcor$exposure = unlist(BMIcor$exposure)
BMIcor$outcome = unlist(BMIcor$outcome)
```



## Write estimates to file

```{r}
f = paste0(project_results_dir, "tables/possible_confounders_on_BMI_20240918.txt")
# write.table(BMIcor, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = TRUE)

BMIcor = read.table(f, header = TRUE, sep = "\t", as.is = TRUE)
```

## BMI is associated with what ? P < 0.05/43

```{r}
BMIcor %>% 
  filter(BMIcor$devexp_exposure_P < 0.05/43) %>% 
  arrange(devexp_exposure_P) %>% dplyr::select(exposure, n, devexp_by_exposure, beta, P) |>
  knitr::kable() %>% kableExtra::kable_classic()
```

## BMI is associated with what at P < 0.05 & > 0.05/43

```{r}
BMIcor %>% 
  filter(BMIcor$devexp_exposure_P < 0.05 & BMIcor$devexp_exposure_P > 0.05/43) %>% 
  arrange(devexp_exposure_P) %>% dplyr::select(exposure, n, devexp_by_exposure, beta, P) |>
  knitr::kable() %>% kableExtra::kable_classic()
```

## Association of Covariables on BMI_GRS

```{r}
BMIGRScor_v0 = t( sapply(cvs, function(cv){
    lmfit_v3( wdata = mydata,
           outcome = "pulit_bmi",
           exposure = cv,
           covariates = NA,
           family = "gaussian",
           weights = NA,
           rnt_outcome = FALSE)
  }) )

###
BMIGRScor_v0 = as.data.frame(BMIGRScor_v0)
  for(i in 3:ncol(BMIGRScor_v0)){
    BMIGRScor_v0[,i] = as.numeric(BMIGRScor_v0[,i])
  }
BMIGRScor_v0$exposure = unlist(BMIGRScor_v0$exposure)
BMIGRScor_v0$outcome = unlist(BMIGRScor_v0$outcome)
```

## Write estimates to file

```{r}
f = paste0(project_results_dir, "tables/possible_confounders_on_PRS_20240918.txt")
# write.table(BMIGRScor_v0, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = TRUE)

BMIGRScor_v0 = read.table(f, header = TRUE, sep = "\t", as.is = TRUE)
```

## What variables associate with the GRS at P < 0.05/43

```{r}
BMIGRScor_v0 %>% filter(BMIGRScor_v0$devexp_exposure_P < 0.05/43) %>%
  arrange(devexp_exposure_P) %>% dplyr::select(exposure, n, devexp_by_exposure, beta, se, devexp_exposure_P) |>
  knitr::kable() %>% kableExtra::kable_classic()
```


## What variables associate with the GRS at P < 0.05 & P > 0.05/43

```{r}
BMIGRScor_v0 %>% filter(devexp_exposure_P < 0.05 & devexp_exposure_P > 0.05/43) %>%
  arrange(devexp_exposure_P) %>% dplyr::select(exposure, n, devexp_by_exposure, beta, devexp_exposure_P) |>
  knitr::kable() %>% kableExtra::kable_classic()
```


### how many MTs did each of the PGS associated variables assocciate with?

```{r}
v = BMIGRScor_v0 |> filter(BMIGRScor_v0$devexp_exposure_P < 0.05/43) |> pull(exposure)

x = BatchCors |> filter(outcome %in% primary_traits) |>
  group_by(exposure) |> 
  summarize( sig_count = sum(devexp_exposure_P < 0.05/33), 
             avg_dev = mean(devexp_by_exposure),
             min_dev = min(devexp_by_exposure),
             max_dev = max(devexp_by_exposure) ) |>
  arrange( desc(sig_count) ) #|> filter(exposure %in% v)

x %>% knitr::kable() %>% kableExtra::kable_classic()

```


```{r}
BMIGRScor = t( sapply(cvs, function(cv){
    lmfit_v3( wdata = mydata,
           outcome = "pulit_bmi",
           exposure = cv,
           covariates = c("sex","age"),
           family = "gaussian",
           weights = NA,
           rnt_outcome = FALSE)
  }) )

###
BMIGRScor = as.data.frame(BMIGRScor)
  for(i in 3:ncol(BMIGRScor)){
    BMIGRScor[,i] = as.numeric(BMIGRScor[,i])
  }
BMIGRScor$exposure = unlist(BMIGRScor$exposure)
BMIGRScor$outcome = unlist(BMIGRScor$outcome)
```

```{r}
BMIGRScor %>% filter(BMIGRScor$devexp_exposure_P < 0.05/43) %>%
  arrange(devexp_exposure_P) %>%
  knitr::kable() %>% kableExtra::kable_classic()
```



```{r}
cat(paste0("***J01_drug_use on BMI***\n"))
fit = lm(BMI ~ J01_antibiotics, data = mydata)
anova(fit)[1, 1:5]


cat(paste0("\n***J01_drug_use on BMI-GRS***\n"))
fit = lm(pulit_bmi ~ J01_antibiotics, data = mydata)
anova(fit)[1, 1:5]

cat(paste0("\n***J01_drug_use on BMI-GRS corrected for sex and age***\n"))
fit = lm(pulit_bmi ~ sex + age + J01_antibiotics, data = mydata)
anova(fit)[3, 1:5]

```



