---
title: "Population Description"
author: "David Hughes"
date: "3/16/2022"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#### LIBRARIES
library(tidyverse)

#### Parameters
source("parameters/pfile.sh")
source("load.R")
```


## Population Description

```{r}
sum( is.na(study_data$clinical_data$sex) )
table( study_data$clinical_data$sex )
table( study_data$clinical_data$imputed_sex )

```


```{r}
mci = function(x, dig = 2){
  m = mean(x, na.rm = TRUE)
  ci = quantile(x, probs = c(0.025, 0.975), na.rm = TRUE)
  out = paste0( round(m, d = dig), " (", round(ci[1],d=dig), "-", round(ci[2],d=dig), ")" )
  return(out)
}
###
a = study_data$clinical_data %>% 
  summarise( sex = "all", mean_age = mci(age),
              mean_height = mci(Height),
              mean_weight = mci(Weight),
              mean_bmi = mci(BMI),
             mean_hh = mci(house_hold_monthl_net_income, dig = 3)
               ) 
####
bs = study_data$clinical_data %>% 
  group_by(sex) %>% 
  summarise( mean_age = mci(age),
              mean_height = mci(Height),
              mean_weight = mci(Weight),
              mean_bmi = mci(BMI),
             mean_hh = mci(house_hold_monthl_net_income, dig = 3)
               ) 

### combine
x = rbind(a, bs)

## table
x %>%  knitr::kable() %>%
  kableExtra::kable_classic()


```

## Smoking Summary (Never Ever Current) smoking numbers 

```{r}
cat( paste0( "all") )
table(study_data$clinical_data$smoking_never_ever_current)

cat( paste0( "\nfemale") )
w = which(study_data$clinical_data$sex == "female")
table(study_data$clinical_data$smoking_never_ever_current[w])

cat( paste0( "\nmale") )
w = which(study_data$clinical_data$sex == "male")
table(study_data$clinical_data$smoking_never_ever_current[w])

cat( paste0( "\nNA") )
w = which( is.na(study_data$clinical_data$sex )) 
table(study_data$clinical_data$smoking_never_ever_current[w])

```


## Household Education

```{r}
cat( paste0( "all") )
table(study_data$clinical_data$house_hold_monthl_net_income)

cat( paste0( "\nfemale") )
w = which(study_data$clinical_data$sex == "female")
table(study_data$clinical_data$house_hold_monthl_net_income[w])

cat( paste0( "\nmale") )
w = which(study_data$clinical_data$sex == "male")
table(study_data$clinical_data$house_hold_monthl_net_income[w])

cat( paste0( "\nNA") )
w = which( is.na(study_data$clinical_data$sex )) 
table(study_data$clinical_data$house_hold_monthl_net_income[w])

```




# Variance explained in exposure by instrument

## Yengo BMI and Height

```{r}
d = data.frame( id = rownames(study_data$clinical_data) ,
                bmi = study_data$clinical_data$BMI, 
                height = study_data$clinical_data$Height,
                whr = study_data$clinical_data$WHR,
                whradjbmi = study_data$clinical_data$WHRadjBMI,
                sex = study_data$clinical_data$sex, 
                imputed_sex = study_data$clinical_data$imputed_sex,
                age = study_data$clinical_data$age,
                yengo = study_data$grs_data$yengo_BMI_wGRS_info_hwe_filtered,
                pulit = study_data$grs_data$pruit_BMI_wGRS_info_hwe_filtered,
                pulit_f = study_data$grs_data$pruit_BMI_females_wGRS_info_hwe_filtered,
                pulit_m = study_data$grs_data$pruit_BMI_males_wGRS_info_hwe_filtered,
                yengo_h = study_data$grs_data$yengo_height_wGRS_info_hwe_filtered,
                pulit_whr = study_data$grs_data$pruit_WHR_wGRS_info_hwe_filtered,
                pulit_whradjbmi = study_data$grs_data$pruit_whradjbmi_wGRS_info_hwe_filtered,
                pc1 = study_data$clinical_data$PC1,
                pc2 = study_data$clinical_data$PC2,
                pc3 = study_data$clinical_data$PC3,
                pc4 = study_data$clinical_data$PC4 )
rownames(d) = d$id

```


```{r}

cat(paste0("BMI in both sexes\n"))
fit = lm( bmi ~ yengo, data = d )
s = summary(fit); s$r.squared

cat(paste0("Height in both sexes\n"))
fit = lm( height ~ yengo, data = d )
s = summary(fit); s$r.squared


```

##  BMI multivariable eta-squared

```{r}
fit = lm( bmi ~ sex + age + yengo, data = d )
a = anova(fit)
ss = a[,2]; names(ss) = rownames(a)
## eta-squared
ss/sum(ss)

```

## R2 on BMI residuals

```{r}
fit = lm( bmi ~ sex + age , data = d )
res = residuals(fit)
##
m = match(names(res), d$id )
fit2 = lm( res ~ d$yengo[m] )
s = summary(fit2); s$r.squared

```



## Pulit BMI

```{r}
f = which(study_data$clinical_data$imputed_sex == "female")
m = which(study_data$clinical_data$imputed_sex == "male")
###
cat(paste0("BMI in both sexes\n"))
fit = lm( bmi ~ pulit, data = d )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in females\n"))
fit = lm( bmi ~ pulit_f, data = d[f,] )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in males\n"))
fit = lm( bmi ~ pulit_m, data = d[m,] )
s = summary(fit); s$r.squared

```

##  BMI multivariable eta-squared

```{r}
fit = lm( bmi ~ sex + age + age^2 + pulit, data = d )

a = anova(fit)
ss = a[,2]; names(ss) = rownames(a)
## eta-squared
ss/sum(ss)

```

## R2 on BMI residuals

```{r}
fit = lm( bmi ~ sex + age + age^2 , data = d )
res = rntransform( residuals(fit) )
##
m = match(names(res), d$id )
fit2 = lm( res ~ d$pulit[m] )
s = summary(fit2); s$r.squared

```


## Pulit WHR

```{r}
f = which(study_data$clinical_data$imputed_sex == "female")
m = which(study_data$clinical_data$imputed_sex == "male")

###
cat(paste0("BMI in both sexes\n"))
fit = lm( study_data$clinical_data$WHR ~ study_data$grs_data$pruit_WHR_wGRS_info_hwe_filtered )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in females\n"))
fit = lm( study_data$clinical_data$WHR[f] ~ study_data$grs_data$pruit_WHR_females_wGRS_info_hwe_filtered[f] )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in males\n"))
fit = lm( study_data$clinical_data$WHR[m] ~ study_data$grs_data$pruit_WHR_males_wGRS_info_hwe_filtered[m] )
s = summary(fit); s$r.squared

```


## Pulit WHRadjBMI

```{r}
f = which(study_data$clinical_data$imputed_sex == "female")
m = which(study_data$clinical_data$imputed_sex == "male")

###
cat(paste0("BMI in both sexes\n"))
fit = lm( study_data$clinical_data$WHRadjBMI ~ study_data$grs_data$pruit_whradjbmi_GRS_info_hwe_filtered )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in females\n"))
fit = lm( study_data$clinical_data$WHRadjBMI[f] ~ study_data$grs_data$pruit_whradjbmi_females_wGRS_info_hwe_filtered[f] )
s = summary(fit); s$r.squared

cat(paste0("\nBMI in males\n"))
fit = lm( study_data$clinical_data$WHRadjBMI[m] ~ study_data$grs_data$pruit_whradjbmi_males_wGRS_info_hwe_filtered[m] )
s = summary(fit); s$r.squared

```

## Correlation between the Yengo and Pulit BMI PGS score

```{r}
cor.test(study_data$grs_data$yengo_BMI_wGRS, study_data$grs_data$pruit_BMI_wGRS )
```

## INFO and HWE filtered PGS

```{r}
cor.test(study_data$grs_data$yengo_BMI_wGRS_info_hwe_filtered, study_data$grs_data$pruit_BMI_wGRS_info_hwe_filtered )
```


```{r, figwidth = 8, fig.height = 7}
mt_data %>% ggplot(aes(x = MDS1, y = MDS2, color = EnterotypeClass)) +
  geom_point(size = 4, alpha = 0.8) +
  scale_color_brewer(palette = "Set1")
  
```

