---
title: "Effect Estimates"
author: "David Hughes"
date: "3/16/2022"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ivreg)

source("parameters/pfile.sh")
source("load.R")

```

## Build an Average DIETARY Consumption Data Set

```{r}
w = grep("average_consumption", colnames(study_data$clinical_data))
acv = colnames(study_data$clinical_data)[w]
##
AvgConVar = sapply(acv, function(v){
  n = gsub("_average_consumption_last_week","", v)
  x = study_data$clinical_data[, v]
  x[x == "Between 5 and 10"] = 6
  x[x == "More than 10"] = 7
  x[x == "Not applicable"] = 0
  return(as.numeric(x) )
})
###
colnames(AvgConVar) = gsub("_average_consumption_last_week","", colnames(AvgConVar) )

```


```{r}
## Define the Microbial Trait Data
mt_data = study_data$study_mt_data
```

## Build the Enterotype binary traits

```{r}
#######################################
## Build the Enterotype binary traits
#######################################
mt_data$Enterotype_Bact1 = mt_data$EnterotypeClass
mt_data$Enterotype_Bact1[mt_data$Enterotype_Bact1 == "Bact1"] = "0"
mt_data$Enterotype_Bact1[mt_data$Enterotype_Bact1 != "0"] = "1"
###
mt_data$Enterotype_Bact2 = mt_data$EnterotypeClass
mt_data$Enterotype_Bact2[mt_data$Enterotype_Bact2 == "Bact2"] = "0"
mt_data$Enterotype_Bact2[mt_data$Enterotype_Bact2 != "0"] = "1"
##
mt_data$Enterotype_Prev = mt_data$EnterotypeClass
mt_data$Enterotype_Prev[mt_data$Enterotype_Prev == "Prev"] = "0"
mt_data$Enterotype_Prev[mt_data$Enterotype_Prev != "0"] = "1"
##
mt_data$Enterotype_Rum = mt_data$EnterotypeClass
mt_data$Enterotype_Rum[mt_data$Enterotype_Rum == "Rum"] = "0"
mt_data$Enterotype_Rum[mt_data$Enterotype_Rum != "0"] = "1"
```

## Build the working data frame

```{r}

## Clinical data
clin_data = study_data$clinical_data[, c("age","sex","BMI", "Height",
                                         "hip_circumference", "waist_circumference",  
                                         "WHR", "WHRadjBMI",
                                         "imputed_sex","imputed_bmi","imputed_height",
                                         "imputed_hip_circumference","imputed_waist_circumference",
                                         "imputed_WHR","imputed_WHRadjBMI",
                                         "PC1", "PC2", "PC3", "PC4",
                                         "smoking_never_ever_current", "house_hold_monthl_net_income",
                                         "is_followingadiet"
                                         )]

## GRS data
grs_data = study_data$grs_data

## Batch Data
batch_data = study_data$batch_data
for(i in 1:8){ batch_data[,i] = as.factor(batch_data[,i])  }

## working data
mydata = cbind( clin_data, grs_data, batch_data, AvgConVar, mt_data )
```

## Identify those who are taking antibiotics

```{r}
w = grep("S01",colnames( study_data$clinical_data) )
S01_drug_use = apply(study_data$clinical_data[,w], 1, function(x){sum(x, na.rm = TRUE)})
sum( S01_drug_use > 0 )

w = grep("D07", colnames( study_data$clinical_data) )
D07_drug_use = apply(study_data$clinical_data[,w], 1, function(x){sum(x, na.rm = TRUE)})
sum( D07_drug_use > 0 )


w = grep("J01", colnames( study_data$clinical_data) )
J01_drug_use = apply(study_data$clinical_data[,w], 1, function(x){sum(x, na.rm = TRUE)})
sum( J01_drug_use > 0 )
J01_drug_use[J01_drug_use>0] = 1

mydata$J01_drug_use = as.factor(J01_drug_use)

## Does BMI associate with Antibiotic use? 
fit = glm( J01_drug_use ~ imputed_sex * BMI , data = mydata,  family = "binomial")
summary(fit)$coef
```

## Identify possible influential batch variables

```{r, warning=FALSE, error=FALSE, message=FALSE}
source("functions/lmfit_v2.R")
## COVARS
cvs = c("Aliquote.date", "drilled.cut", "Alliquoted.by", "Date.of.extraction", "Extraction.done.by",
        "PCR.Plate", "Position", "LIBRARY_DATE")
##
test = t( sapply(MTs, function(mt){
  lmfit_v2( wdata = wdata,
          outcome = mt,
          exposure = "total_read_count",
          covariates = cvs,
          weights = NA,
          rnt_outcome = TRUE,
          typeIIanova = TRUE)
}) )
###  
test = data.frame(test)
for(i in 3:ncol(test)){ test[,i] = as.numeric(test[,i]) }
```


```{r}
w = grep("_P", colnames(test))
w = w[-c(1:4)]
( sig_count = apply(test[,w], 2, function(x){ sum(x< 0.05) }) )

```


```{r}
w = grep("etasq_", colnames(test) )

(mean_eta = apply(test[,w], 2, function(x){ mean(x, na.rm = TRUE) }) )
```


```{r}
batch_covariate_4_sensitivity = c( "drilled.cut", "Aliquote.date", "Date.of.extraction", 
                           "PCR.Plate", "Position", "LIBRARY_DATE",
                           "total_read_count" )
```


## Set up analysis matrix

```{r}
analysis_matrix = data.frame(  
  exposure = c( rep("BMI",3) , rep("WHR",3), rep("WHRadjBMI",3) ),
  instrument = c("pruit_BMI_wGRS_info_hwe_filtered", "pruit_BMI_females_wGRS_info_hwe_filtered", "pruit_BMI_males_wGRS_info_hwe_filtered",
                 "pruit_WHR_wGRS_info_hwe_filtered", "pruit_WHR_females_wGRS_info_hwe_filtered", "pruit_WHR_males_wGRS_info_hwe_filtered",
                 "pruit_whradjbmi_wGRS_info_hwe_filtered", "pruit_whradjbmi_females_wGRS_info_hwe_filtered", "pruit_whradjbmi_males_wGRS_info_hwe_filtered" 
                 ),
  sex = rep(c("all","female","male"),3),
  analysis_names = c("BMI","BMI_females","BMI_males",
                     "WHR","WHR_females","WHR_males",
                     "WHRadjBMI","WHRadjBMI_females","WHRadjBMI_males")
  )
```

# Derive Estimates

```{r}
Derived_Estimates = list()

####################################################
## (I) Iterate over each row of the analysis matrix
####################################################
for(i in 1:nrow(analysis_matrix)){
  ## Parameters
  expo = analysis_matrix[i, "exposure"]
  inst = analysis_matrix[i, "instrument"]
  sex = analysis_matrix[i, "sex"]
  anal = analysis_matrix[i, "analysis_names"]
  ### Run an analysis for PA then AB traits
  mod = c("binomial","gaussian")
  suffix = c("PA", "AB")
  
  ####################################################
  ## (II) Define the data frame to run
  ####################################################
  data2run = mydata
  
  if(sex == "female"){
    w = which(mydata$imputed_sex == "female")
    data2run = mydata[w,]
  }
    
  if(sex == "male"){
    w = which(mydata$imputed_sex == "male")
    data2run = mydata[w,]
  }
  
  ####################################################
  ## (III) Define the covariates to run
  ####################################################
  covars2run = c("imputed_sex","age")
  
  if(sex != "all"){ covars2run = "age" }
  
  ####################################################
  ## (IV) Run PA then AB traits
  ####################################################
  Dout = c()
  for(j in 1:2 ){
    ##########################
    ## (IV.a) Define Traits 2 be run
    ##########################
    if(suffix[j] == "PA"){
      w = grep("_PA", colnames(mydata))
      mts = colnames(mydata)[w] 
      ## add enterotype traits to PA run
      ent = c("Enterotype_Bact1","Enterotype_Bact2","Enterotype_Prev","Enterotype_Rum")
      mts = c(ent, mts)
    } else{
      w = grep("_AB", colnames(mydata) )
      mts = colnames(mydata)[w]
      ## add diversity traits to AB run
      div = c("MDS1","MDS2","Div_NumberGenera","Div_Shannon","Div_Chao1")
      mts = c(div, mts)
    }
    ##########################
    ## (IV.b) Define MODEL 2 be run
    ##########################
    model2run = mod[j]
    ##########################
    ## (IV.c) run the analysis
    ##########################
    Eout = t( sapply(mts, function(mt){
    ivtoolsfit( wdata = mydata,
              outcome = mt,
              exposure = expo,
              instrument = inst,
              covariates = covars2run,
              outcome_model_family = model2run,
              exposure_model_family = "gaussian",
              weights = NA,
              rnt_outcome = TRUE)
      } ) )
    ## (IV.d) Define as a data frame
    Eout = as.data.frame(Eout)
    ## (IV.e) Make numeric
    for(k in 4:ncol(Eout)){  Eout[,k] = as.numeric(Eout[,k])  }
    ## (IV.f) Define trait
    Eout$trait_type = suffix[j]
    ## (IV.g) redefine column 14 as it can be a tvalue or zvalue depending on logistic or linear regression
    colnames(Eout)[14] = "oe_tzvalue"
    ## (IV.h) place PA and AB data together 
    Dout = rbind(Dout, Eout)
  }
  ## (IV.i) Add a data run flg to the data frame
  Dout$analysis_name = anal
    
  ##########################
  ## V) Return data
  ##########################
  Derived_Estimates[[anal]] = Dout
}

names(Derived_Estimates)
```

## Save the estimates as an Rdata file

```{r}
f = paste0(project_results_dir, "Derived_Estimates.Rdata")
save(Derived_Estimates, file = f)
```

## Merge BMI the short data together

```{r}
bmi_merged = rbind(Derived_Estimates$BMI, Derived_Estimates$BMI_females, Derived_Estimates$BMI_males)
f = paste0(project_results_dir, "tables/bmi_obs_mr_estimates_v0.2.txt")
write.table(bmi_merged, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = TRUE)
```


## Average Variance Explained in exposure by instrument

```{r}
unlist( lapply(Derived_Estimates, function(x){ mean(x$ei_eta_sq)  }) )
```


## BMI MTs causally associated with BMI

```{r}
Derived_Estimates$BMI %>% filter(MR_P < 0.05/44) %>% 
  arrange(MR_P) %>% 
  dplyr::select( MR_n,  MR_beta, MR_se, MR_P, MR_Wald_F, trait_type, analysis_name) %>%
  knitr::kable() %>% kableExtra::kable_classic()
```



## Sensitivity Analysis

```{r}
####################################################
## (I) Define the covariates to run
####################################################
covars2run = c( "drilled.cut", 
                "imputed_sex","age", 
                "smoking_never_ever_current","house_hold_monthl_net_income",
                "J01_drug_use" , 
                "is_followingadiet", "sugar_free_soda" )
  
####################################################
## (II) Run PA then AB traits
####################################################
### Run an analysis for PA then AB traits
mod = c("binomial","gaussian")
suffix = c("PA", "AB")
  
Dout = c()
for(j in 1:2 ){
  ##########################
  ## (III.a) Define Traits 2 be run
  ##########################
  if(suffix[j] == "PA"){
    w = grep("_PA", colnames(mydata))
    mts = colnames(mydata)[w]  
    ## add enterotype traits to PA run
    ent = c("Enterotype_Bact1","Enterotype_Bact2","Enterotype_Prev","Enterotype_Rum")
    mts = c(ent, mts)
  } else{
    w = grep("_AB", colnames(mydata) )
    mts = colnames(mydata)[w]
    ## add diversity traits to AB run
    div = c("MDS1","MDS2","Div_NumberGenera","Div_Shannon","Div_Chao1")
    mts = c(div, mts)
  }
  ##########################
  ## (III.b) Define MODEL 2 be run
  ##########################
  model2run = mod[j]
  ##########################
  ## (III.c) run the analysis
  ##########################
  Eout = t( sapply(mts, function(mt){
  ivtoolsfit( wdata = mydata,
            outcome = mt,
            exposure = "BMI",
            instrument = "pruit_BMI_wGRS_info_hwe_filtered",
            covariates = covars2run,
            outcome_model_family = model2run,
            exposure_model_family = "gaussian",
            weights = NA,
            rnt_outcome = TRUE)
    } ) )
  ## (III.d) Define as a data frame
  Eout = as.data.frame(Eout)
  ## (III.e) Make numeric
  for(k in 4:ncol(Eout)){  Eout[,k] = as.numeric(Eout[,k])  }
  ## (III.f) Define trait
  Eout$trait_type = suffix[j]
  ## (III.g) redefine column 14 as it can be a tvalue or zvalue depending on logistic or linear regression
  colnames(Eout)[14] = "oe_tzvalue"
  ## (III.h) place PA and AB data together 
  Dout = rbind(Dout, Eout)
}

BMI_sensitivity = Dout
BMI_sensitivity$analysis_name = "BMI_sensitivity"
```


## Save the estimates as an Rdata file

```{r}
f = paste0(project_results_dir, "BMI_sensitivity_v0.1.Rdata")
save(BMI_sensitivity, file = f)
```



```{r}
BMI_sensitivity %>% filter(MR_P < 0.05/44) %>% 
  arrange(MR_P) %>% 
  dplyr::select( MR_n,  MR_beta, MR_se, MR_P, MR_Wald_F, trait_type, analysis_name) %>%
  knitr::kable() %>% kableExtra::kable_classic()
```





## Alternative Phenotype definition

```{r}
fit = lm(BMI ~ imputed_sex + age + age^2, data = mydata)
bmi_res = rntransform( residuals(fit) )
##
m = match(rownames(mydata), names(bmi_res))
mydata$bmi_res = bmi_res[m]
```


```{r}
####################################################
## (I) Define the covariates to run
####################################################
covars2run = NA
  
####################################################
## (II) Run PA then AB traits
####################################################
### Run an analysis for PA then AB traits
mod = c("binomial","gaussian")
suffix = c("PA", "AB")
  
Dout = c()
for(j in 1:2 ){
  ##########################
  ## (III.a) Define Traits 2 be run
  ##########################
  if(suffix[j] == "PA"){
    w = grep("_PA", colnames(mydata))
    mts = colnames(mydata)[w]  
    ## add enterotype traits to PA run
    ent = c("Enterotype_Bact1","Enterotype_Bact2","Enterotype_Prev","Enterotype_Rum")
    mts = c(ent, mts)
  } else{
    w = grep("_AB", colnames(mydata) )
    mts = colnames(mydata)[w]
    ## add diversity traits to AB run
    div = c("MDS1","MDS2","Div_NumberGenera","Div_Shannon","Div_Chao1")
    mts = c(div, mts)
  }
  ##########################
  ## (III.b) Define MODEL 2 be run
  ##########################
  model2run = mod[j]
  ##########################
  ## (III.c) run the analysis
  ##########################
  Eout = t( sapply(mts, function(mt){
  ivtoolsfit( wdata = mydata,
            outcome = mt,
            exposure = "bmi_res",
            instrument = "pruit_BMI_wGRS_info_hwe_filtered",
            covariates = NA,
            outcome_model_family = model2run,
            exposure_model_family = "gaussian",
            weights = NA,
            rnt_outcome = TRUE)
    } ) )
  ## (III.d) Define as a data frame
  Eout = as.data.frame(Eout)
  ## (III.e) Make numeric
  for(k in 4:ncol(Eout)){  Eout[,k] = as.numeric(Eout[,k])  }
  ## (III.f) Define trait
  Eout$trait_type = suffix[j]
  ## (III.g) redefine column 14 as it can be a tvalue or zvalue depending on logistic or linear regression
  colnames(Eout)[14] = "oe_tzvalue"
  ## (III.h) place PA and AB data together 
  Dout = rbind(Dout, Eout)
}

BMI_alt_BMIexposure = Dout
BMI_alt_BMIexposure$analysis_name = "BMIalt"
```


## Save the estimates as an Rdata file

```{r}
f = paste0(project_results_dir, "BMI_Pulit_Residuals_v0.1.Rdata")
save(BMI_alt_BMIexposure, file = f)
```



```{r}
BMI_alt_BMIexposure %>% filter(MR_P < 0.05/44) %>% 
  arrange(MR_P) %>% 
  dplyr::select( MR_n,  MR_beta, MR_se, MR_P, MR_Wald_F, trait_type, analysis_name) %>%
  knitr::kable() %>% kableExtra::kable_classic()
```


