---
title: "Define Study Microbial Traits"
author: "David Hughes"
date: "3/16/2022"
output: html_document
---

How shall we handle the taxonomic count filtering and the building of the MTs (microbial traits) ?


```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)

#### LIBRARIES
library(tidyverse)

#### Parameters
source("parameters/pfile.sh")
# source("load.R")

## Load Rdata file from script step03
f = paste0(project_data_dir, "study_data_20240911_v1.1.Rdata")
load( f )

## source the needed function file
source( "functions/build_mts.R" )
```

## Define initial study microbial traits data

```{r}

############################################
## Limit the microbial traits to those with
##  -- 1. at least 20 non-zero individuals
##  ***  NO STEP 2 FILTER  ***
##  -- 2. a zero-truncated mean count of 25 OR a minimum of 500 (5%) reads in at least 1 individual
############################################
## all taxa in data set
all_mt_taxa = colnames(study_data$micro_data)[12:509]

## Build the MT data
MTdata = build_mts( wdata = study_data$micro_data[, all_mt_taxa], 
                      PA_min_obs = 250,   ## The minimum number of PRESENT of ABSCENT Observations to build a PA trait
                      trunc_AB_present_count = 1000, ## The minimum number of present observations to build a trunc_AB
                      trunc_AB_absence_count = 250, ## The minimum number of absent (0's) observations to build a trunc_AB
                      trunc_AB_mean = 20,  ## The minimum zero-truncated mean to build a trunc_AB
                      AB_zero_present_proportion = 0.75, ## The maximum proportion of zero data to build an AB trait
                      AB_mean_of_zero_trunc_data = 20 ## The minimum zero-truncated mean to build an AB trait
)
MTdata = as.data.frame(MTdata)

dim(MTdata)
```


## Build Ratio Traits

```{r}
## Identify the AB traits
w = grep("_AB", colnames(MTdata))
ids = colnames(MTdata)[w]
## Do not keep the truncated AB traits
w = grep("trunc_AB", ids)
ids = ids[-w]

## Keep only those AB traits with a median count of 25
med_count = apply(MTdata[, ids], 2, function(x){ median(x, na.rm = TRUE) })
w = which(med_count >= 25)
ids = ids[w]

## Build a Spearman's Correlation matrix
Cmat = cor(MTdata[, ids], method = "spearman", use = "pairwise.complete.obs")

## Which trait pairs inversely correlate ? 
l = apply(Cmat, 1, function(x){which(x < -0.2)})
w =  which( lapply(l, length) >0 )
l = l[w]
##
vals = list()
for(i in names(l)){
  r = l[[i]]
  est = Cmat[i, r]
  names(est) = names(l[[i]])
  vals[[i]] = est
}
```

## Identify trait pairs that correlate (pos and neg)

```{r}
x = 1
ratio_trait_pairs = c()
for(i in 1:ncol(Cmat)){
  for(j in x:nrow(Cmat)){
    # cat(paste0(i, " ", j,"\n"))
    
    ## Identify trait pairs that inversely correlate
    #if( Cmat[i,j] <= -0.2){
    
    if( abs(Cmat[i,j]) >= 0.2 & abs(Cmat[i,j]) <= 0.9 ){
      o = c(rownames(Cmat)[i],  colnames(Cmat)[j], Cmat[i,j] )
      ratio_trait_pairs = rbind(ratio_trait_pairs, o)
    }
  }
  x = x + 1
}
ratio_trait_pairs = as.data.frame(ratio_trait_pairs)
colnames(ratio_trait_pairs) = c("trait_1","trait_2","Spearman_rho")
```

## Define the ratios

```{r}

new_ratio_trait_names = c()
MT_ratios = c()
for(i in 1:nrow(ratio_trait_pairs)){
  tps = as.character( ratio_trait_pairs[i,1:2] )
  n1 = tps[1]; n1 = strsplit(n1, split = "_")[[1]]
  w = grep("AB", n1); w = c(w-1, w); n1 = paste(n1[-w], collapse = "_")
  ##
  n2 = tps[2]; n2 = strsplit(n2, split = "_")[[1]]
  w = grep("AB", n2); w = c(w-1, w); n2 = paste(n2[-w], collapse = "_")
  ##
  n = paste0( n1, "|", n2)
  newtrait = (MTdata[,tps[1]] +1) / (MTdata[,tps[2]] +1)
  ##
  new_ratio_trait_names = c(new_ratio_trait_names, n)
  MT_ratios = cbind(MT_ratios, newtrait)
}
colnames(MT_ratios) = new_ratio_trait_names
colnames(MT_ratios) = paste0(colnames(MT_ratios), "_RA")

dim(MT_ratios)
```

## Redefine ratio_trait_pairs

```{r}
ratio_trait_pairs$ratio_trait_id = new_ratio_trait_names

f = paste0(project_results_dir, "tables/ratio_trait_pairs_20240917.txt")
write.table(ratio_trait_pairs, file = f, col.names = TRUE, row.names = FALSE, sep = "\t", quote = FALSE)
```


## Redefine the MTdata

```{r}
MTdata = cbind(MTdata, MT_ratios)
dim(MTdata)
```


## Correlation Structure

```{r , warning=FALSE, message=FALSE, error=FALSE}
## Build a Spearman's correlation matrix
Cmat = cor(MTdata, method = "sp", use = "pairwise.complete.obs")
```


```{r, fig.width = 20, fig.height = 20}
M = Cmat
M[is.na(M)] = 0

corrplot::corrplot(M, method = 'color', order = 'hclust',  diag = FALSE, tl.cex = 0.3, tl.col = 'black')
```


```{r}
f = paste0(project_results_dir, "figures/mt_corplot_20240917.pdf")
# f = paste0(project_results_dir, "figures/mt_corplot_v2_with_ratios.pdf")
pdf(file = f, width = 25, height = 25)
corrplot::corrplot(M, method = 'color', order = 'hclust',  diag = FALSE, tl.cex = 0.25, tl.col = 'black')
dev.off()

```

## Identify the redundant, higher order taxa (MTs)

```{r}
Dmat = as.dist(1-abs(M))
nj = hclust(Dmat)
k = cutree(nj, h = 0.05)
nrow(M)
length(unique(k))
```

## Identify Primary Analysis taxa list

```{r}
TAXAORDER = c("K","P","C","O","F","G")

## Identify K group with more than 1 MT
# K_groups = names(  which( table(k) > 1 ) )
K_groups = unique(k)

## Identify the primary trait list
primary_traits = c()

## Run a Loops
for(i in K_groups){
  ## extract MT names
  mt_names = names( which(k == i) )
  
  ## STEP 1: screen clusters that have (trunc)_AB, traits - we will always take an abundance trait over a PA or a RA trait
  if( length( grep("AB", mt_names) ) > 0  ){
    ## Let's limit the our list of possible traits to just the (trunc)AB traits
    w = grep("AB", mt_names)
    mt_names = mt_names[w]
  }
  
  ## STEP 2: screen for trunc_AB and AB traits
  if( length( grep("trunc_AB", mt_names) ) > 0 & length( grep("trunc_AB", mt_names) ) != length(mt_names)  ){
    ## remove the trunc_AB traits
    w = grep("trunc_AB", mt_names)
    mt_names = mt_names[-w]
  }
  
  ## Now just take the last trait in the vector as it should be the lowest order trait
  primary_traits = c(primary_traits, mt_names[length(mt_names)])
  
}
```


## Concatenate all of the traits names for each k.

```{r}
K_traits = c()
for(i in K_groups){
  mt_names = names( which(k == i) )
  if(length(mt_names) > 1){
    x = paste(mt_names, collapse = " ; ")   
  } else {
    x = mt_names
  }
  
  K_traits = c(K_traits, x)
}
```

```{r}
K_df = data.frame(primary_trait = primary_traits, 
                  number_of_traits_in_cluster = as.numeric( table(k) ), 
                  traits_in_cluster = K_traits)
f = paste0(project_results_dir, "tables/Trait_Clusters_Primary_Trait_20240917.txt")
write.table(K_df, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = TRUE)
```



## Identify taxa to remove

```{r}
# TAXAORDER = c("K","P","C","O","F","G")
# 
# ## Identify K group with more than 1 MT
# K_groups = names(  which( table(k) > 1 ) )
# ## Extract those MTs that ARE NOT the lowest taxonomic order
# taxa_2_remove = c()
# ##
# for(i in K_groups){
#   ## extract MT names
#   mt_names = names( which(k == i) )
#   ## screen for trunc_AB and AB traits
#   if( length( grep("trunc_AB", mt_names) ) > 0 & length( grep("trunc_AB", mt_names) ) != length(mt_names)  ){
#     ## take the trunc_AB taxa
#     q = grep("trunc_AB", mt_names)
#     trunc_AB_mts = mt_names[q]
#     AB_mts = mt_names[-q]
#     ##########################
#     ## Process the trunc_AB taxa
#     ##########################
#     if(length(trunc_AB_mts) > 1){
#       to = substring(trunc_AB_mts, 1,1)
#       m = na.omit( match(TAXAORDER, to) )
#       r = m[-length(m)]
#       taxa_2_remove = c(taxa_2_remove, trunc_AB_mts[r] )
#     }
#     ##########################
#     ## Process the AB taxa
#     ##########################
#     if(length(AB_mts) > 1) {
#       to = substring(AB_mts, 1,1)
#       m = na.omit( match(TAXAORDER, to) )
#       r = m[-length(m)]
#       taxa_2_remove = c(taxa_2_remove, AB_mts[r] )
#     }
#   
#     ### If the mt_names do not include both trunc_AB and AB taxa names
#     ### Process here:
#   } else {
#     if( length( grep("_RA", mt_names) ) > 0 ) {
#       ## Keep the top order and remove the lower ones
#       r = mt_names[2:length(mt_names)]
#       taxa_2_remove = c(taxa_2_remove, r )
#     } else {
#       ## extract taxa order
#       to = substring(mt_names, 1,1)
#       m = na.omit( match(TAXAORDER, to) )
#       r = m[-length(m)]
#       taxa_2_remove = c(taxa_2_remove, mt_names[r] )
#     }
#   }
#   
#   
# }
```

## Redefine Study_micro_data

```{r}
study_micro_data = cbind( study_data$micro_data[c("enterotype", "EnterotypeClass", 
                                                  "Enterotype_Bact1","Enterotype_Bact2","Enterotype_Prev","Enterotype_Rum",
                                                  "MDS1", "MDS2", 
                                                  "Div_NumberGenera","Div_Shannon","Div_Chao1")] , 
                          MTdata )
```


```{r}
## Define a STUDY MICROBIOAL TRAIT Data Frame
study_data$study_micro_data = study_micro_data
study_data$primary_traits = primary_traits

```


```{r}
#################################
## RE-WRITE The project data frame
#################################
# "study_data_v1.2.Rdata" written April 3rd 2023
n = paste0( project_data_dir , "study_data_20240917_v1.2.Rdata")
save(study_data, file = n)

```


## Correlation Structure of STUDY MT DATA set

```{r , warning=FALSE, message=FALSE, error=FALSE}
## remove the enterotype data
temp = study_data$study_micro_data[,-c(1:2)]
## find the primary traits, while also keeping the diversity metrics
w = c(1:9, which( colnames(temp) %in% primary_traits) )
## redefine temp
temp = temp[w,]
for(i in 1:ncol(temp)){
  temp[, i] = as.numeric( temp[, i] )
}

## Build a Spearman's correlation matrix -->
Cmat = cor( temp, method = "sp", use = "pairwise.complete.obs")

```


```{r, fig.width = 20, fig.height = 20}
M = Cmat
M[is.na(M)] = 0 

corrplot::corrplot(M, method = 'color', order = 'hclust',  diag = FALSE, tl.cex = 0.3, tl.col = 'black') 
```


```{r}
f = paste0(project_results_dir, "figures/mt_corplot_primarytraits_20240917.pdf")
pdf(file = f, width = 20, height = 20)
corrplot::corrplot(M, method = 'color', order = 'hclust',  diag = FALSE, tl.cex = 0.25, tl.col = 'black')
dev.off()

```


