---
title: "Effect Estimates"
author: "David Hughes"
date: "3/16/2022"
output: html_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ivreg)

source("parameters/pfile.sh")
source("load.R")

```


```{r}
## Define the Microbial Trait Data
mt_data = study_data$study_mt_data
```

## Build the Enterotype binary traits

```{r}
#######################################
## Build the Enterotype binary traits
#######################################
mt_data$Enterotype_Bact1 = mt_data$EnterotypeClass
mt_data$Enterotype_Bact1[mt_data$Enterotype_Bact1 == "Bact1"] = "0"
mt_data$Enterotype_Bact1[mt_data$Enterotype_Bact1 != "0"] = "1"
###
mt_data$Enterotype_Bact2 = mt_data$EnterotypeClass
mt_data$Enterotype_Bact2[mt_data$Enterotype_Bact2 == "Bact2"] = "0"
mt_data$Enterotype_Bact2[mt_data$Enterotype_Bact2 != "0"] = "1"
##
mt_data$Enterotype_Prev = mt_data$EnterotypeClass
mt_data$Enterotype_Prev[mt_data$Enterotype_Prev == "Prev"] = "0"
mt_data$Enterotype_Prev[mt_data$Enterotype_Prev != "0"] = "1"
##
mt_data$Enterotype_Rum = mt_data$EnterotypeClass
mt_data$Enterotype_Rum[mt_data$Enterotype_Rum == "Rum"] = "0"
mt_data$Enterotype_Rum[mt_data$Enterotype_Rum != "0"] = "1"
```


## Build an Average DIETARY Consumption Data Set

```{r}
w = grep("average_consumption", colnames(study_data$clinical_data))
acv = colnames(study_data$clinical_data)[w]
##
AvgConVar = sapply(acv, function(v){
  n = gsub("_average_consumption_last_week","", v)
  x = study_data$clinical_data[, v]
  x[x == "Between 5 and 10"] = 6
  x[x == "More than 10"] = 7
  x[x == "Not applicable"] = 0
  return(as.numeric(x) )
})
###
colnames(AvgConVar) = gsub("_average_consumption_last_week","", colnames(AvgConVar) )

```

## Identify those who are taking antibiotics

```{r}
w = grep("S01",colnames( study_data$clinical_data) )
S01_drug_use = apply(study_data$clinical_data[,w], 1, function(x){sum(x, na.rm = TRUE)})
sum( S01_drug_use > 0 )

w = grep("D07", colnames( study_data$clinical_data) )
D07_drug_use = apply(study_data$clinical_data[,w], 1, function(x){sum(x, na.rm = TRUE)})
sum( D07_drug_use > 0 )


w = grep("J01", colnames( study_data$clinical_data) )
J01_drug_use = apply(study_data$clinical_data[,w], 1, function(x){sum(x, na.rm = TRUE)})
sum( J01_drug_use > 0 )
J01_drug_use[J01_drug_use>0] = 1

study_data$clinical_data$J01_drug_use = as.factor(J01_drug_use)

## Does BMI associate with Antibiotic use? 
fit = glm( J01_drug_use ~ imputed_sex * BMI , data = study_data$clinical_data,  family = "binomial")
summary(fit)$coef
```

## Build the working data frame

```{r}

## Clinical data
clin_data = study_data$clinical_data[, c("age","sex","BMI", "Height",
                                         "hip_circumference", "waist_circumference",  
                                         "WHR", "WHRadjBMI",
                                         "imputed_sex","imputed_bmi","imputed_height",
                                         "imputed_hip_circumference","imputed_waist_circumference",
                                         "imputed_WHR","imputed_WHRadjBMI",
                                         "PC1", "PC2", "PC3", "PC4",
                                         "smoking_never_ever_current", "house_hold_monthl_net_income",
                                         "is_followingadiet", "J01_drug_use"
                                         )]

## GRS data
grs_data = study_data$grs_data

## Batch Data
batch_data = study_data$batch_data
for(i in 1:8){ batch_data[,i] = as.factor(batch_data[,i])  }

## working data
mydata = cbind( clin_data, grs_data, batch_data, AvgConVar, mt_data )
```

## Read in the results

```{r}
f = paste0(project_results_dir, "Derived_Estimates.Rdata")
load(file = f)
##
f = paste0(project_results_dir, "BMI_sensitivity_v0.1.Rdata")
load(file = f)
##
f = paste0(project_results_dir, "BMI_Pulit_Residuals_v0.1.Rdata")
load(file = f)
```


```{r}
# f = paste0(project_results_dir, "tables/bmi_obs_mr_estimates_v0.2.txt")
# bmi_merged = read.table(f, header = TRUE, sep = "\t", as.is = TRUE)
```


## Average Variance Explained in exposure by instrument

```{r}
unlist( lapply(Derived_Estimates, function(x){ mean(x$ei_eta_sq)  }) )
```


## BMI MTs causally associated with BMI

```{r}
tout = Derived_Estimates$BMI %>% filter(MR_P < 0.05/44) %>% 
  arrange(MR_P) %>% 
  dplyr::select( trait_type, oe_n, oe_beta, oe_se, oe_P, oe_exposure_eta_sq,  MR_beta, MR_se, MR_P, MR_Wald_F, MR_dhat_r2)
  
  
tout %>% knitr::kable() %>% kableExtra::kable_classic()
```

```{r}
f = paste0(project_results_dir, "tables/table1_topresults.txt")
write.table(tout, file = f, row.names = TRUE, col.names = TRUE, sep = "\t", quote = TRUE)
```



## Forest plot for top BMI results

```{r, fig.width = 10, fig.height = 8}
BMI_long = ivtoolsfit_2_longformat(data = Derived_Estimates$BMI, 
                                        pop_id = "BMI", alpha = 0.05/44 )


## Define analysis factor order
BMI_long$analysis = factor(BMI_long$analysis, levels = c("tsls_BMI", "obs_BMI")[2:1] )

## Define Colors
x = RColorBrewer::brewer.pal(8, "Blues")[8]
y = RColorBrewer::brewer.pal(8, "Reds")[8]
pcol = c(y,x)


## MAKE PLOT
plot = dh_forrest_plot(data = BMI_long, 
                           alpha = 0.05/44, 
                           analysis_type = "tsls_BMI",
                           arrange_by = "pval",
                           title = NA, 
                           covariates = NA, 
                           confounders = NA,
                           outcome_label = "outcome : MTs",
                           exposure_label = "exposure : BMI",
                           filter_data = TRUE,
                           column_4_x = "analysis",
                           column_4_color = "analysis",
                           col_count = 1,
                           color_choices = pcol)

plot
```


```{r}
f = paste0(project_results_dir, "figures/bmi_forrest_top_results.pdf")
pdf(f, width = 10, height = 8)
plot
dev.off()
```


## Forest plot for nominal BMI results

```{r, fig.width = 20, fig.height = 12}
BMI_long = ivtoolsfit_2_longformat(data = Derived_Estimates$BMI, 
                                        pop_id = "BMI", alpha = 0.05/44 )


## Define analysis factor order
BMI_long$analysis = factor(BMI_long$analysis, levels = c("tsls_BMI", "obs_BMI")[2:1] )

## Define Colors
x = RColorBrewer::brewer.pal(8, "Blues")[8]
y = RColorBrewer::brewer.pal(8, "Reds")[8]
pcol = c(y,x)


## MAKE PLOT
plot = dh_forrest_plot(data = BMI_long, 
                           alpha = 0.05, 
                           analysis_type = "tsls_BMI",
                           arrange_by = "beta",
                           title = NA, 
                           covariates = NA, 
                           confounders = NA,
                           outcome_label = "outcome : MTs",
                           exposure_label = "exposure : BMI",
                           filter_data = TRUE,
                           column_4_x = "analysis",
                           column_4_color = "analysis",
                           col_count = 2,
                           color_choices = pcol)

plot
```


```{r}
f = paste0(project_results_dir, "figures/bmi_forrest_nominal_results.pdf")
pdf(f, width = 20, height = 12)
plot
dev.off()
```



## INCLUDE BOTH SEXES

## Convert BMI (males and females) to long format and merge

```{r}
##########################
## convert 2 long format
##########################
BMI_long = ivtoolsfit_2_longformat(data = Derived_Estimates$BMI, 
                                        pop_id = "BMI", alpha = 0.05/44 )
BMI_females_long = ivtoolsfit_2_longformat(data = Derived_Estimates$BMI_females, 
                                        pop_id = "BMI_females", alpha = 0.05/44 )
BMI_males_long = ivtoolsfit_2_longformat(data = Derived_Estimates$BMI_males, 
                                        pop_id = "BMI_males", alpha = 0.05/44 )
##########################
## Merge data sets
##########################
BMI_long_all = rbind(BMI_long, BMI_females_long, BMI_males_long)
```

## Forest plot with both sexes included

```{r, fig.width = 10, fig.height = 12}
## Define analysis factor order
BMI_long_all$analysis = factor(BMI_long_all$analysis, levels = c("tsls_BMI", "tsls_BMI_females", "tsls_BMI_males",
                                                         "obs_BMI", "obs_BMI_females", "obs_BMI_males" )[6:1] )

## Define Colors
x = RColorBrewer::brewer.pal(8, "Blues")[8:6]
y = RColorBrewer::brewer.pal(8, "Reds")[8:6]
pcol = c(y,x)


## MAKE PLOT
plot = dh_forrest_plot(data = BMI_long_all, 
                           alpha = 0.05/44, 
                           analysis_type = "tsls_BMI",
                           arrange_by = "pval",
                           title = NA, 
                           covariates = NA, 
                           confounders = NA,
                           outcome_label = "outcome : MTs",
                           exposure_label = "exposure : BMI",
                           filter_data = TRUE,
                           column_4_x = "analysis",
                           column_4_color = "analysis",
                           col_count = 1,
                           color_choices = pcol)

plot
```

```{r}
f = paste0(project_results_dir, "figures/bmi_forest_top_results_both_sexes.pdf")
pdf(f, width = 10, height = 12)
plot
dev.off()
```


## Comparison between Observational and MR effect estimates

-- for PA and AB traits 
-- BMI is the exposure
-- amongst all individuals

```{r, fig.width = 10, fig.height = 10}
BMI_results = Derived_Estimates$BMI

## BMI OBservational Significance
BMI_results$oe_sig = "Obs P > 0.05"
w = which(BMI_results$oe_P < 0.05)
if(length(w)>0){ BMI_results$oe_sig[w] = "Obs P < 0.05" }
w = which(BMI_results$oe_P < 0.05/46)
if(length(w)>0){ BMI_results$oe_sig[w] = paste0("Obs P < ", round(0.05/46, d = 4) ) }
BMI_results$oe_sig = factor(BMI_results$oe_sig, 
                      levels = c( "Obs P > 0.05", 
                                  "Obs P < 0.05", 
                                  paste0("Obs P < ", round(0.05/46, d = 4) ) ) )
## BMI MR Significance
BMI_results$MR_sig = "MR P > 0.05"
w = which(BMI_results$MR_P < 0.05)
if(length(w)>0){ BMI_results$MR_sig[w] = "MR P < 0.05" }
w = which(BMI_results$MR_P < 0.05/46)
if(length(w)>0){ BMI_results$MR_sig[w] = paste0("MR P < ", round(0.05/46, d = 4) ) }
BMI_results$MR_sig = factor(BMI_results$MR_sig, 
                      levels = c( "MR P > 0.05", 
                                  "MR P < 0.05", 
                                  paste0("MR P < ", round(0.05/46, d = 4) ) ) )


#################
## PLOT
#################
plot = BMI_results %>% ggplot( aes(x = oe_beta, y = MR_beta) ) +
  geom_smooth(formula = y~x, method = loess, color = "black") + 
  geom_smooth(formula = y~x, method = lm, color = "grey50") +
  geom_point( aes( shape = oe_sig, fill = MR_sig ), size = 4, alpha = .80 ) +
  scale_shape_manual(values = c(21,22,23)) +
  scale_fill_brewer(palette = "Set1") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  theme_bw() +
  guides(fill = guide_legend( override.aes = list (shape=c(24) ) ) ) +
  labs( fill = "MR significance", shape = "Obs. significance" ,
        x = "Observational effect estimates", y = "MR effect estimates") +
  # ggrepel::geom_text_repel( data = BMI_results %>% filter(MR_P < 0.05/46) , 
  #                 aes(label=MR_outcome), 
  #                 min.segment.length = 0,
  #                 box.padding = 0.4,
  #                 nudge_x = 0.015,
  #                 nudge_y = -0.01,
  #                 label.size = 3) +
  facet_wrap(. ~ trait_type, nrow = 2, scales = "free") 


plot
```


```{r}
f = paste0(project_results_dir, "figures/obs_vs_mr.pdf")
pdf(f, width = 10, height = 8)
plot
dev.off()
```


```{r}
cor.test(BMI_results$oe_beta, BMI_results$MR_beta)
```

```{r}
w = which(BMI_results$trait_type == "AB")
cor.test(BMI_results$oe_beta[w], BMI_results$MR_beta[w])
```

```{r}
w = which(BMI_results$trait_type == "PA")
cor.test(BMI_results$oe_beta[w], BMI_results$MR_beta[w])

```



## Diversity traits with NO Rank Normal Transformation

```{r}
## Add diversity traits
div = c("MDS1","MDS2","Div_NumberGenera","Div_Shannon","Div_Chao1")
mts = c(div)

## Derive Estiamtes
div_iv = t( sapply(mts, function(mt){
  ivtoolsfit( wdata = mydata,
            outcome = mt,
            exposure = "BMI",
            instrument = "pruit_BMI_wGRS_info_hwe_filtered",
            covariates = c( "imputed_sex","age"),
            outcome_model_family = "gaussian",
            exposure_model_family = "gaussian",
            weights = NA,
            rnt_outcome = FALSE)
}) )

####
div_iv = as.data.frame(div_iv)
###
for(i in 4:ncol(div_iv)){ div_iv[,i] = as.numeric(div_iv[,i])  }
div_iv$trait_type = "AB"
colnames(div_iv)[14] = "oe_ABt_PAz_val"
```


```{r}
div_iv %>% 
  dplyr::select( oe_beta, oe_se, oe_P, MR_n,  MR_beta, MR_se, MR_P, MR_Wald_F) %>%
  knitr::kable() %>% kableExtra::kable_classic()
```

# Confounders


## Is the BMI GRS correlated with household income?

```{r}
fit = glm(pruit_BMI_wGRS_info_hwe_filtered ~ house_hold_monthl_net_income, data = mydata)
summary(fit)$coef
```

## Is the BMI GRS correlated with smoking? 

```{r}
mydata$smoking_never_ever_current = factor(mydata$smoking_never_ever_current, 
                                           levels = c("never", "ever", "current"  ) )

fit = lm(pruit_BMI_wGRS_info_hwe_filtered ~ as.numeric( smoking_never_ever_current ), data = mydata)
summary(fit)$coef
#anova(fit)
```


## Is the BMI GRS correlated with 16S batch variables? 

```{r}
bvars = colnames(mydata)[68:74]

Batch_Var_PGS_Association = sapply(bvars, function(v){
  form = formula( paste0("pruit_BMI_wGRS_info_hwe_filtered ~ ", as.factor( v ) ) )
  #form = formula( paste0("BMI ~ ", as.factor( v ) ) )
  fit = lm(form, data = mydata)
  a = anova(fit)
  p = a[1,5]
  return(p)
})

Batch_Var_PGS_Association

# fit = lm(pruit_BMI_wGRS_info_hwe_filtered ~ Date.of.extraction , data = mydata)
# summary(fit)$coef
# anova(fit)
```

# Are Dietary variables confounders ??

## Dietary variables

```{r}
Dvars = c( "is_followingadiet", "is_gaining_weight_unintentionally", colnames(study_data$clinical_data)[1012:1102] )

## Some Other Variables of interest
OtherVars = c("sleeping_hours_per_day", "moderate_activities", "exercising_habits", "daily_seated_work_hours")
```

## Is the BMI GRS correlated with any dietary variables? 

```{r}
###
DVars_2_use = c("sleeping_hours_per_day", "moderate_activities", "exercising_habits", "daily_seated_work_hours", 
                "is_followingadiet", "is_gaining_weight_unintentionally", "is_vegetarian", "is_vegan" )
##
TempData = cbind(mydata, study_data$clinical_data[, DVars_2_use], AvgConVar)
##
DVars_2_use = c(DVars_2_use, colnames(AvgConVar) )
##
Diet_Var_PGS_Association = t( sapply(DVars_2_use, function(v){
  form = formula( paste0("pruit_BMI_wGRS_info_hwe_filtered ~  as.numeric(",v,")" ) )
  fit = lm(form, data = TempData)
  s = summary(fit)
  coef = s$coef[2, c(1,2,4)]
  r2 = s$r.squared
  a = anova(fit)
  Fstat = a[1,4]
  out = c(coef, r2, Fstat)
  names(out) = c("Beta","SE","P","R2","F")
  return(out)
}) )

Diet_Var_PGS_Association %>% knitr::kable() %>% kableExtra::kable_classic()

```



## Evaluation of the BMI Sensitivity Effect Estimates

```{r}
BMI_sensitivity %>% filter(MR_P < 0.05/44) %>% 
  arrange(MR_P) %>% 
  dplyr::select( MR_n,  MR_beta, MR_se, MR_P, MR_Wald_F, trait_type, analysis_name) %>%
  knitr::kable() %>% kableExtra::kable_classic()
```


## Convert BMI (v0 and sensitivity) to long format and merge

```{r}
##########################
## convert 2 long format
##########################
BMI_long = ivtoolsfit_2_longformat(data = Derived_Estimates$BMI, 
                                        pop_id = "BMI", alpha = 0.05/44 )

BMI_sens_long = ivtoolsfit_2_longformat(data = BMI_sensitivity, 
                                        pop_id = "BMI_sensitivity", alpha = 0.05/44 )

##########################
## Merge data sets
##########################
new_BMI_long = rbind(BMI_long, BMI_sens_long )
```

## Forest plot with both sexes included

```{r, fig.width = 10, fig.height = 12}
## Define analysis factor order
new_BMI_long$analysis = factor(new_BMI_long$analysis, levels = c("tsls_BMI", "tsls_BMI_sensitivity",
                                                         "obs_BMI", "obs_BMI_sensitivity"  )[4:1] )

## Define Colors
x = RColorBrewer::brewer.pal(8, "Blues")[8:7]
y = RColorBrewer::brewer.pal(8, "Reds")[8:7]
pcol = c(y,x)


## MAKE PLOT
plot = dh_forrest_plot(data = new_BMI_long, 
                           alpha = 0.05/44, 
                           analysis_type = "tsls_BMI",
                           arrange_by = "pval",
                           title = NA, 
                           covariates = NA, 
                           confounders = NA,
                           outcome_label = "outcome : MTs",
                           exposure_label = "exposure : BMI",
                           filter_data = TRUE,
                           column_4_x = "analysis",
                           column_4_color = "analysis",
                           col_count = 1,
                           color_choices = pcol)

plot
```


```{r}
f = paste0(project_results_dir, "figures/bmi_forest_top_results_sensitivity.pdf")
pdf(f, width = 10, height = 12)
plot
dev.off()
```

```{r}
##########################
## convert 2 long format
##########################
BMI_long = ivtoolsfit_2_longformat(data = Derived_Estimates$BMI, 
                                        pop_id = "BMI", alpha = 0.05/44 )
BMI_females_long = ivtoolsfit_2_longformat(data = Derived_Estimates$BMI_females, 
                                        pop_id = "BMI_females", alpha = 0.05/44 )
BMI_males_long = ivtoolsfit_2_longformat(data = Derived_Estimates$BMI_males, 
                                        pop_id = "BMI_males", alpha = 0.05/44 )

BMI_sens_long = ivtoolsfit_2_longformat(data = BMI_sensitivity, 
                                        pop_id = "BMI_sensitivity", alpha = 0.05/44 )


##########################
## Merge data sets
##########################
BMI_long_all = rbind(BMI_long, BMI_sens_long, BMI_females_long, BMI_males_long)
```

## Forest plot with both sexes included

```{r, fig.width = 10, fig.height = 12}
## Define analysis factor order
BMI_long_all$analysis = factor(BMI_long_all$analysis, levels = c("tsls_BMI", "tsls_BMI_sensitivity", "tsls_BMI_females", "tsls_BMI_males",
                                                         "obs_BMI", "obs_BMI_sensitivity", "obs_BMI_females", "obs_BMI_males" )[8:1] )


## Define Colors
x = RColorBrewer::brewer.pal(8, "Blues")[c(8:7,5:4)]
# x2 = RColorBrewer::brewer.pal(8, "Greens")[8:7]
y = RColorBrewer::brewer.pal(8, "Reds")[c(8:7, 5:4)]
# y2 = RColorBrewer::brewer.pal(8, "Greens")[6:5]
#pcol = c(y, y2, x ,x2)
pcol = c(y, x)

## MAKE PLOT
plot = dh_forrest_plot(data = BMI_long_all, 
                           alpha = 0.05/44, 
                           analysis_type = "tsls_BMI",
                           arrange_by = "pval",
                           title = NA, 
                           covariates = NA, 
                           confounders = NA,
                           outcome_label = "outcome : MTs",
                           exposure_label = "exposure : BMI",
                           filter_data = TRUE,
                           column_4_x = "analysis",
                           column_4_color = "analysis",
                           col_count = 1,
                           color_choices = pcol)

plot


```


```{r}
f = paste0(project_results_dir, "figures/bmi_forest_top_results_sex_and_sensitivity.pdf")
pdf(f, width = 10, height = 12)
plot
dev.off()
```






## Comparison between Observational and MR effect estimates using the alternative (residual) BMI exposure

-- for PA and AB traits 
-- BMI is the exposure
-- amongst all individuals

```{r, fig.width = 10, fig.height = 10}
BMI_results = BMI_alt_BMIexposure

## BMI OBservational Significance
BMI_results$oe_sig = "Obs P > 0.05"
w = which(BMI_results$oe_P < 0.05)
if(length(w)>0){ BMI_results$oe_sig[w] = "Obs P < 0.05" }
w = which(BMI_results$oe_P < 0.05/46)
if(length(w)>0){ BMI_results$oe_sig[w] = paste0("Obs P < ", round(0.05/46, d = 4) ) }
BMI_results$oe_sig = factor(BMI_results$oe_sig, 
                      levels = c( "Obs P > 0.05", 
                                  "Obs P < 0.05", 
                                  paste0("Obs P < ", round(0.05/46, d = 4) ) ) )
## BMI MR Significance
BMI_results$MR_sig = "MR P > 0.05"
w = which(BMI_results$MR_P < 0.05)
if(length(w)>0){ BMI_results$MR_sig[w] = "MR P < 0.05" }
w = which(BMI_results$MR_P < 0.05/46)
if(length(w)>0){ BMI_results$MR_sig[w] = paste0("MR P < ", round(0.05/46, d = 4) ) }
BMI_results$MR_sig = factor(BMI_results$MR_sig, 
                      levels = c( "MR P > 0.05", 
                                  "MR P < 0.05", 
                                  paste0("MR P < ", round(0.05/46, d = 4) ) ) )


#################
## PLOT
#################
plot = BMI_results %>% ggplot( aes(x = oe_beta, y = MR_beta) ) +
  geom_smooth(formula = y~x, method = loess, color = "black") + 
  geom_smooth(formula = y~x, method = lm, color = "grey50") +
  geom_point( aes( shape = oe_sig, fill = MR_sig ), size = 4, alpha = .80 ) +
  scale_shape_manual(values = c(21,22,23)) +
  scale_fill_brewer(palette = "Set1") +
  geom_hline(yintercept = 0, color = "black", linetype = "dashed") +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  theme_bw() +
  guides(fill = guide_legend( override.aes = list (shape=c(24) ) ) ) +
  labs( fill = "MR significance", shape = "Obs. significance" ,
        x = "Observational effect estimates", y = "MR effect estimates") +
  # ggrepel::geom_text_repel( data = BMI_results %>% filter(MR_P < 0.05/46) , 
  #                 aes(label=MR_outcome), 
  #                 min.segment.length = 0,
  #                 box.padding = 0.4,
  #                 nudge_x = 0.015,
  #                 nudge_y = -0.01,
  #                 label.size = 3) +
  facet_wrap(. ~ trait_type, nrow = 2, scales = "free") 


plot
```

```{r}
cor.test(BMI_results$oe_beta, BMI_results$MR_beta)
```

```{r}
w = which(BMI_results$trait_type == "AB")
cor.test(BMI_results$oe_beta[w], BMI_results$MR_beta[w])
```

```{r}
w = which(BMI_results$trait_type == "PA")
cor.test(BMI_results$oe_beta[w], BMI_results$MR_beta[w])

```



